---
title: "exercise1"
author: "Alexande Van Roijen"
date: "January 9, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Problem 1

```{r cars}
set.seed(123)
smallSample = (rnorm(10,0,1))
hist(smallSample)
qqnorm(smallSample)
abline(0,1)
```
surprisingly, this does look rather normal! or at least the qqplot says so. The histogram does not look normal, more uniform if anything.

```{r prob1}
for(i in 1:20){
  smallSample = (rnorm(10,0,1))
  par(mfrow=c(1,2))
  hist(smallSample)
  qqnorm(smallSample)
  abline(0,1)
}
```
After generating 20 seperate instances, we see we were a bit lucky with the qqplot and it doesnt really come up to snuff.

Problem 2
```{r prob2}
x = seq(-4,4,.01)
plot(dt(x,3),type='l')
lines(dnorm(x,0,1),type='l')
```
as we can see, there is a heavier tail and smaller and flatter peak for the t distribution

Problem 3
```{r prob3}
tSample = rt(10,3)
hist(tSample)
qqnorm(tSample)
abline(0,1)
```
Surpirisngly, it looks pretty normal via the qqplot, but not in the histogram, just like before

Problem 4
```{r prob4}
n=100
smallSample = (rnorm(n,0,1))
tSample = rt(n,3)
hist(smallSample)
qqnorm(smallSample)
abline(0,1)
hist(tSample)
qqnorm(tSample)
abline(0,1)
```
We seem to have found a normal looking noramlly generated distrib, but not very true to the actual curve is the t distributed sample.

Problem 5
```{r prob5}
sampleMeans = numeric(200)
sampleSize = 10
for(i in 1:200)
{
  sampleMeans[i]=mean(rnorm(sampleSize))
}
hist(sampleMeans)
print(mean(sampleMeans))
print((var(sampleMeans)))
print(sqrt(var(sampleMeans)))
print(1/sqrt(sampleSize))
```

Problem 6
```{r prob6}
sampleMeans = numeric(200)
sampleSize = 40
for(i in 1:200)
{
  sampleMeans[i]=mean(rnorm(sampleSize))
}
hist(sampleMeans)
print(mean(sampleMeans))
print((var(sampleMeans)))
print(sqrt(var(sampleMeans)))
print(1/sqrt(sampleSize))
```

Problem 7
```{r prob7}
print(1/.01)
```
by dividing the true variance by the standard error we desire, we see that 100 sample size is needed theoretically to achieve it.

Problem 8
```{r prob8}
bernoulSample = rbinom(10,5,0.1)
print((.1*.9)/(0.01**2))
```
In the above, we knew the standard error, which is the SD of the sample proportion, is 0.01, and then knew the true variance follows p*q. We simple invert, square, and solve for n
Problem 9
```{r prob9}
expMeans = numeric(200)
for(i in 1:200)
{
  expMeans[i] = mean(rexp(10,3))
}
hist(expMeans)
qqnorm(expMeans)
qqline(expMeans)
```

problem 10
```{r prob10}
print(mean(expMeans))
print(var(expMeans))
print(sqrt(var(expMeans)))
print("versus theoreticals")
print((1/3))
print((1/9)/10)
print(sqrt((1/9)/10))

```

They are pretty close!

problem 11
```{r prob11}
print("problem number 5 repeated")
sampleMeans = numeric(200)
sampleSize = 40
for(i in 1:200)
{
  sampleMeans[i]=mean(rnorm(sampleSize))
}
hist(sampleMeans)
print(mean(sampleMeans))
print((var(sampleMeans)))
print("versus theoreticals")
print(sqrt(var(sampleMeans)))
print(1/sqrt(sampleSize))
```
problem 12
```{r prob12}
sampleMeans = numeric(200)
sampleSize = 10
for(i in 1:200)
{
  sampleMeans[i]=mean(rbinom(sampleSize,1,0.1))
}
hist(sampleMeans)
sampleMeans = numeric(200)
sampleSize = 40
for(i in 1:200)
{
  sampleMeans[i]=mean(rbinom(n=sampleSize,size=1,prob=0.1))
}
hist(sampleMeans)

sampleMeans = numeric(200)
sampleSize = 100
for(i in 1:200)
{
  sampleMeans[i]=mean(rbinom(n=sampleSize,size=1,prob=0.1))
}
hist(sampleMeans)
```

Problem 13
```{r prob13}
sampleMeans = numeric(200)
sampleSize = 10
for(i in 1:200)
{
  sampleMeans[i]=mean(rbinom(sampleSize,1,0.5))
}
hist(sampleMeans)
sampleMeans = numeric(200)
sampleSize = 40
for(i in 1:200)
{
  sampleMeans[i]=mean(rbinom(n=sampleSize,size=1,prob=0.5))
}
hist(sampleMeans)

sampleMeans = numeric(200)
sampleSize = 100
for(i in 1:200)
{
  sampleMeans[i]=mean(rbinom(n=sampleSize,size=1,prob=0.5))
}
hist(sampleMeans)
```

The n goes down with a p of .5. The reason being that our variance has decreased, which means our n has also decreased as well to achieve a tighet interval. or at least thats how i undersand it


Problem 14
``` {r prob14}

```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
