---
title: "Exercise 5"
author: "Alexander Van Roijen"
date: "February 13, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Problem 1

```{r prob1.1}
wineData = read.csv("wine.csv")
alcData = wineData[c("alc","quality")]
aTest = aov(quality~factor(alc),data=wineData)
print(summary(aTest))
```

It appears there is a difference in the average quality of the wine depending on the level of alcohol used. Considering the control of the experiment and the randomization, we can take a further step and state that there is some causal effect on the quality of the wine based on the alcohol level.

```{r prob1.2}
phTest = aov(quality~factor(ph),data=wineData)
#IT IS IMPORTANT TO USE FACTOR HERE IN PARTICULAR
print(summary(phTest))
```

It appears there is no evidence to suggest a difference in mean quality levels given different ph levels.

```{r prob1.3}
almTest = lm(quality~alc,data=wineData)
print(summary(almTest))
plot(y=wineData$quality,x=wineData$alc)
abline(almTest)
```

From the graphic and summary, we can see there is a strong positive linear trend displayed between the three different alcohol levels and rated quality of the wine. This agrees that there is signifigance with the anova test.

```{r prob1.4}
phlmTest = lm(quality~ph,data=wineData)
print(summary(phlmTest))
plot(y=wineData$quality,x=wineData$ph)
abline(phlmTest)
```

Again, we can see that our lm agrees with our ANOVA test and shows little to no signifigant correlation between ph levels and quality.


###Problem 2

```{r prob2.1}
noInterTest = aov(quality~factor(alc)+factor(ph),wineData)
print(summary(noInterTest))
```

They are both significant in this case, as we can tell by te very high p-values. This is due to a variety of reasons, but for this particular example, we can think of it as the noise being dampened as specified by the 2 way anova equation listed in the slides.

```{r prob2.2}
lmNITest = lm(quality ~ alc + ph,wineData)
print(summary(lmNITest))
```

Similar to our anova test, we are seeing significant p values for our coefficients in regards to their impact on quality.

```{r prob2.3}
noInterTest = aov(quality~factor(alc)*factor(ph),wineData)
print(summary(noInterTest))
```


according to this anova test, we can see that they are still both significant, but not significane is seen for their interaction